Using Interactive Graphics to Visualize Evening Grosbeak Movement and Migration Patterns

### 2024-05-09 paper.qmd
actively working through R Shiny results, discussion, and conclusion. Also made one spelling change in the app

@@

<img src="eg_pic.jpeg" alt="Evening Grosbeak image" width="400"/>

@@

## Background

-   Why are we looking at winter spots? What needs to be analyzed?
-   What are the possibilities of Shiny?
-   Write about R

<p><em>Motus</em> is a wildlife tracking system from <em>Birds Canada</em>. It is composed of a network of researchers around the world who are able to contribute to the radio telemetry system. By contributing (tagging birds and setting up radio receiver locations), more researchers are then able to access the growing database of bird tracking information. While one researcher could hypothetically set up their own, small telemetry network to track a local bird, <em>Motus</em> creates the opportunity to connect all of these local sites and broaden data and information exponentially. Collaboration for scientists and bird-watchers across the world has been made easy by <em>Motus</em>.</p>

@@

@@

### Motus R Package

@@

```{r, echo=TRUE, include=TRUE, error=FALSE, warning=FALSE, message=FALSE, results='hide'}
install.packages(c("motus", "motusData"), 
                 repos = c(birdscanada = 'https://birdscanada.r-universe.dev',
                           CRAN = 'https://cloud.r-project.org'))

@@

-   downloading project data and then exporting detections (include data table of initial data file)
-   understanding the variables (descriptions of what they mean, say which ones are actually relevant to this project)

@@

### Exploration of Data

-   initial plots and exploration

@@

<p>There are many benefits to using Shiny to showcase data analyses and visualizations. To start, the interactive ability of a Shiny app can transform user accessibility to data. With these apps, there are many widgets that the creator can include in the UI (that then speak to the server output) that allows a user to choose what they are seeing from the data: slider inputs (for numerical input), text inputs, checkboxes, dropdown menus, file uploaders, etc. With this user power to choose outputs, Shiny apps can be reactive and responsive, updating the outputs of each page in real-time. Another benefit is ease fo deployment for Shiny apps. Instead of code files or papers showcasing research, Shiny apps can be deployed easily with URLS and showcase findings in a more fluid, interactive, and engaging way.</p>

@@

-   screenshots of app
-   visualizations
-   leaflet maps of deployments
-   leaflet maps of winter spots

## Discussion

-   future applications
-   bio/data science student to build off of this starter app and analyze what is actually happening
-   needs more data cleaning (birds in south america should not be showing up)
-   more widgets for timelines (slider input to pick a time period to plot a bird)

@@

>>>>

```{r, echo = FALSE, include=FALSE}
library(shiny)
library(leaflet)
library(dplyr)
library(shinythemes)
library(DT)
library(bslib)
library(lubridate)
library(knitr)
```

```{r, echo = FALSE, include=FALSE}
eg_df <- readRDS("df_eg.RDS")

eg_df <- eg_df %>%
  mutate(motusTagDepID = paste(motusTagID, tagDeployID, sep = "."))
eg_df <- eg_df %>%
  relocate(motusTagDepID, .after = motusTagID)
```

Methods
- motus in R
- Shiny (show default app, super simple examples of each thing (not using bird data)) (basically a Shiny tutorial)
- leaflets (big tools that we will be using for this specific app)
- (shiny live (you can start with static images))

Results
- code

@@

<!-- ![Evening Grosbeak image](eg_pic.jpeg){width = "100"} -->
<!-- {.lightbox} -->
<img src="eg_pic.jpeg" alt="Evening Grosbeak image" width="400"/>

@@

@@

### Motus
<p><em>Motus</em> is a wildlife tracking system from <em>Birds Canada</em>. It is composed of a network of researchers around the world who are able to contribute to the radio telemetry system. By contributing (tagging birds and setting up radio receiver locations), more researchers are then able to access the growing database of bird tracking information. While one researcher could hypothetically set up their own, small telemetry network to track a local bird, <em>Motus</em> creates the opportunity to connect all of these local sites and broaden data and information exponentially. Collaboration for scientists and bird-watchers across the world has been made easy by <em>Motus</em>.</p>

@@

#### Motus R Package

@@

```{r, echo=TRUE, include=TRUE, error=FALSE, warning=FALSE, message=FALSE, results='hide'}
#| code-fold: show
#| code-summary: "Click to show code"

# Use these commented lines to install the packages
# install.packages(c("motus", "motusData"), 
#                  repos = c(birdscanada = 'https://birdscanada.r-universe.dev',
#                            CRAN = 'https://cloud.r-project.org'))

@@

<p> Once connected to the project, one can work with the data while connected to <em>motus</em>, but this will take lots of time and computer power. It is easier to download the data until the most recent detection, and then save as a .RDS file so that one can work faster with a local dataset as a flat dataframe. </p>

<p>The data has many variables with explanations taken from <a>https://motuswts.github.io/motus/articles/03-accessing-data.html</a></p>

<details>
<summary><b>Variable Descriptions</b></summary>

| Variable | Description |
|----|----------------------------|
| activity | data related to radio activity for each hour period (hourBin) at each antenna, including a count of the number of short runs used in helping identify false detections. |
| admInfo | internal table used to keep track of your the motus package used to create your motus file, and the data version. |
| antDeps | metadata related to antenna deployments, e.g., deployment height, angle, antenna type. |
| batchRuns | metadata for runIDs and associated batchIDs |
| batches | detection data for a given receiver and boot number. |
| filters | metadata related to user created filters associated with the specified receiver. |
| gps | metadata related to Geographic Positioning System (GPS) position of receiver. |
| hits | detection data at the level of individual hits. |
| meta | metadata related to the project and datatype (tags vs. receivers) that are included in the .motus file |
| nodeData | data related to nodes by batchID and time (ts) |
| nodeDeps | metadata related to nodes |
| projAmbig | metadata related to what projects have ambiguous tag detections |
| projs | metadata related to projects, e.g., project name, principal investigator. |
| pulseCounts | number of radio pulses measured on each antenna over each hour period (hourBin). |
| recvDeps | metadata related to receiver deployments, e.g., deployment date, location, receiver characteristics. |
| recvs | metadata related to receiver serial number and associated Motus deviceID |
| runs | detection data associated with a run (continuous detections of a unique tag on a given receiver). |
| runsFilters | a list of runIDs associated with user created filters and assigned probabilities. |
| species | metadata related to species, e.g., unique identifier, scientific name, common name. |
| tagAmbig | metadata related to ambiguous tags, e.g., ambigID and associated motusTagID |
| tagDeps | metadata related to tag deployments, e.g., deployment date, location, and species. |
| tagProp | metadata related to custom deployment properties entered by the principal investigator (e.g. body weight). |
| tags | metadata related to tags, e.g., unique identifier, tag characteristics (e.g., burst interval). |
| allambigs | lists in long-data format each motusTagID (up to 6) associated with each negative ambigID. |
| alltags | provides the full detection data for all tags, and all ambiguous (duplicate) tags, associated with your project. Ambiguous detections are repeated for each motusTagID represented by each ambigID. |
| alltagsGPS | same as alltags but includes GPS latitude, longitude and altitude (much slower to load on large databases). |

</details>

@@

<p>Once the data from the .motus is made into a flat data frame, data can be cleaned. The first step was to trim the exhaustive project data down to just the Evening Grosbeak.</p>

```{r, echo=TRUE, include=TRUE, error=FALSE, warning=FALSE, message=FALSE, results='hide', eval = FALSE}
#| code-fold: show
#| code-summary: "Click to show code"

# filter to only Evening Grosbeak 
df <- tbl.alltags %>%
  filter(speciesID == 20570) %>%   # Evening Grosbeak Species ID
  collect() %>% 
  as.data.frame() %>%    
  mutate(time = as_datetime(ts)) 
# This also changes ts to a date variable instead of numeric (how it was read in)
```

<p>From the data, it is also important to be able to grab a unique bird. A unique tag (either <em>motusTagID</em> or <em>tagDeployID</em>) is not quite capable of this. This is because, a unique <em>motusTagID</em> is not always a unique bird; occasionally a tag can be redeployed onto a different bird. Also a <em>tagDeployID</em> is not always unique when you are focusing on a specific deployment of a bird in your project. So, use a combination of these.</p>

```{r, echo=TRUE, include=TRUE, error=FALSE, warning=FALSE, message=FALSE, results='hide', eval = FALSE}
#| code-fold: show
#| code-summary: "Click to show code"
df <- df_alltags %>%
  mutate(motusTagDepID = paste(motusTagID, tagDeployID, sep = "."))
```

<p>Here is how the new variable would look:</p>

```{r, echo=FALSE}
eg_df %>%
  select(motusTagID, tagDeployID, motusTagDepID, recvDeployLat, recvDeployLon, ts) %>%
  head() %>%
  distinct(motusTagDepID, .keep_all = TRUE) %>%
  kable()
```

<p>Now, we can grab the data for any single bird's deployment with <em>motusTagDepID</em>. </p>

@@

<p>There are many benefits to using Shiny to showcase data analyses and visualizations. To start, the interactive ability of a Shiny app can transform user accessibility to data. With these apps, there are many widgets that the creator can include in the UI (that then speak to the server output) that allows a user to choose what they are seeing from the data: slider inputs (for numerical input), text inputs, checkboxes, dropdown menus, file uploaders, etc. With this user power to choose outputs, Shiny apps can be reactive and responsive, updating the outputs of each page in real-time. Another benefit is ease of deployment for Shiny apps. Instead of code files or papers portraying research, Shiny apps can be deployed easily with URLS and showcase findings in a more fluid, interactive, and engaging way.</p>

<p>The following is the starting structure for a  Shiny App in R:</p>

```{r, eval=FALSE}
#| code-fold: show
#| code-summary: "Click to show Shiny App default code"

#
# This is a Shiny web application. You can run the application by clicking
# the 'Run App' button above.
#
# Find out more about building applications with Shiny here:
#
#    http://shiny.rstudio.com/
#

library(shiny)

# Define UI for application that draws a histogram
ui <- fluidPage(

    # Application title
    titlePanel("Old Faithful Geyser Data"),

    # Sidebar with a slider input for number of bins 
    sidebarLayout(
        sidebarPanel(
            sliderInput("bins",
                        "Number of bins:",
                        min = 1,
                        max = 50,
                        value = 30)
        ),

        # Show a plot of the generated distribution
        mainPanel(
           plotOutput("distPlot")
        )
    )
)

# Define server logic required to draw a histogram
server <- function(input, output) {

    output$distPlot <- renderPlot({
        # generate bins based on input$bins from ui.R
        x    <- faithful[, 2]
        bins <- seq(min(x), max(x), length.out = input$bins + 1)

        # draw the histogram with the specified number of bins
        hist(x, breaks = bins, col = 'darkgray', border = 'white',
             xlab = 'Waiting time to next eruption (in mins)',
             main = 'Histogram of waiting times')
    })
}

# Run the application 
shinyApp(ui = ui, server = server)
```

![Default Shiny App](default_shiny.png){.lightbox}

<p> We can see here that this app has an option in the UI for <em>sliderInput</em> where the user can control the number of bins in the histogram. As the user interacts with this widget, the number of bins on the histogram is changed by the server's response and generation of output for this input.</p>

<p>Next is an example of a dropdown menu (using the IRIS dataset: a popular example dataset about iris dimensions).</p>

```{r, eval=FALSE}
#| code-fold: show
#| code-summary: "Click to see dropdown examples"

library(shiny)

ui <- fluidPage(
  titlePanel("Iris"),
  sidebarLayout(
    sidebarPanel(
      selectInput("x_var", "X-axis Variable", choices = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")),
      selectInput("y_var", "Y-axis Variable", choices = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")),
    ),
    mainPanel(
      plotOutput("histogram")
    )
  )
)

server <- function(input, output) {

  output$histogram <- renderPlot({
    summary(iris[[input$x_var]])
    scatter.smooth(iris[[input$x_var]], main = paste("Scatterplot of", input$x_var))
  })
}

shinyApp(ui = ui, server = server)
```

![Iris dropdown menus Shiny App](iris_dropdown.png){.lightbox gallery = "iris"}
![Dropdown menu](dropdown.png){.lightbox gallery = "iris"}

@@

-   screenshots of app
-   visualizations
-   leaflet maps of deployments
-   leaflet maps of winter spots
- you can talk obout what is seen a bit, don't have to go into detail about whether or not the data makes sense (just is it summarizing what we intend it to summarize, is it actually doing what I claim its doing. The experts actually analyze it.)

<p>The interactive Evening Grosbeak platform was made using Shiny. From the navigation bar, the user can choose from "<em>Deployments and Tracking</em>," "<em>Winter Roosting</em>," or "<em>About</em>." Each navigation page includes multiple tabs for exploration.</p>

![Interactive Evening Grosbeak Platform](deployments.png){.lightbox}

### <em>Deployments and Tracking</em>
<p>Upon launching the app, the user is met with the <em>Deployments and Tracking</em> page. </p>

### <em>Winter Roosting</em>

### <em>About</em>

## Discussion
-  obstacles 
-   future applications
-   bio/data science student to build off of this starter app and analyze what is actually happening
-   needs more data cleaning (birds in south america should not be showing up) (find a map that shows bird in South America and say what you can learn from these shiny app results to further clean the data)
-   summer spots
-   more widgets for timelines (slider input to pick a time period to plot a bird)

@@

I would like to thank Dr. Ivan Ramler for his support and guidance throughout this project. <br>
I would also like to thank Dr. Susan Willson for the motivation behind this project.
